Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-02-24 paper/main.tex
small progress in section 3

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

These observations motivate our study of a technique reducing the
number of acceptance marks without altering the structure of the
automaton. This paper presents a technique heavily based on
\emph{quantified Boolean formulas (QBF)}. For a given TELA and
parameters $C,K$, it produces a QBF formula which is satisfiable if
and only if there exists an automaton with the same structure, $K$
acceptance marks, an acceptance formula in disjunctive normal form
with $C$ clauses, and the same set of accepting runs as the original
automaton. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of
two simpler formulas such that their satisfiability implies the
existence of an automaton with the same structure, $K$ acceptance
marks, and the same set of accepting runs, but not vice versa.

@@

%A transition $(p,a,M',q)\in\delta$ is a transition leading from state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$
A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $M(\pi)$ denote set of all acceptance marks that occur
in infinitely many transitions of run $\pi$. Run $\pi$ is
\emph{accepting} iff it satisfies the formula $\varphi$, where $\pi$
satisfies $\Inf m$ iff $m\in M(\pi)$ and $\pi$ satisfies $\Fin m$
iff $m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the
set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. A nonempty set of states $S\subseteq Q$
is called a nontrivial \emph{strongly connected component (SCC)} if
for each $p,q\in S$ there is a path from $p$ to $q$. An SCC S is
\emph{maximal} if there is no SCC $S'$ satisfying $S\subsetneq S'$. In
the rest of this paper, SCC always refers to a maximal SCC. Given an
SCC $S$, $\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$
denotes the set of all transitions between states in $S$. Further, for
each mark $m\in M$, let $\delta_m$ denotes the set of all transitions
marked with $m$, i.e., $\delta_m=\{(p,a,M',q)\in\delta\mid m\in
M'\}$. The sets $\delta_m$ are called \emph{acceptance sets}.  Given
an SCC $S$ and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
$\delta_{S,m}=\delta_S\cap\delta_m$. Finally, a set of transitions
$D\subseteq\delta$ is called a \emph{cycle} is there exists a path
from a state $p$ to the same state containing all transitions of $D$
and no transition outside $D$.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$ without altering its
structure and language. In other words, we change only the acceptance
marks on individual transitions and we replace the acceptance formula
by a new one with less acceptance marks. To ensure that the modified
automaton is equivalent to the original one, we actually construct the
new acceptance formula such that the set of acceptance runs (modulo
changed acceptance marks on transitions) is preserved.

We start with the following simple observation. Definovat $inf(\varphi)$?

>>>>

\newcommand{\Nset}{{\mathbb{N}_0}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
%\def\Inf{\ensuremath{\mathsf{Inf}}}
%\def\Fin{\ensuremath{\mathsf{Fin}}}
\DeclareMathOperator{\Inf}{\mathsf{Inf}}
\DeclareMathOperator{\Fin}{\mathsf{Fin}}
\newcommand{\minf}{\mathit{inf}}
\def\false{\mathit{false}}
\def\true{\mathit{true}}
\newcommand{\mA}{\mathcal{A}}
\newcommand{\mB}{\mathcal{B}}

@@

These observations motivate our study of a technique reducing the
number of acceptance marks without altering the structure of the
automaton. This paper presents a technique heavily based on
\emph{quantified Boolean formulas (QBF)}. For a given TELA and
parameters $k,n$, it produces a QBF formula which is satisfiable if
and only if there exists an automaton with the same structure, $n$
acceptance marks, an acceptance formula in disjunctive normal form
with $k$ clauses, and the same set of accepting runs as the original
automaton. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of
two simpler formulas such that their satisfiability implies the
existence of an automaton with the same structure, $n$ acceptance
marks, and the same set of accepting runs, but not vice versa.

@@

A transition $t=(p,a,M',q)\in\delta$ is the transition leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$. We
use $M(t)$ to refer to its set of acceptance marks $M'$. We extend
this notation to a set $T\subseteq\delta$ of transitions such that
$M(T)=\bigcup_{t\in T}M(t)$ denotes the set of acceptance marks that
appear on transitions in $T$.

\todo{Rozmyslet si, jak dal.}

A run $\pi$ of $\mA$ over an infinite word
$u = u_{0}u_{1}u_{2}\dots \in \Sigma^{\omega}$ is an infinite sequence
of adjacent transitions
$\pi=(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots\in\delta^{\omega}$ where
$q_0=q_I$. Let $\minf(\pi)$ denote the set of transitions that appear
infinitely many times in $\pi$. Further, let $M(\pi)$ denote the set
of acceptance marks that occur in infinitely many transitions of run
$\pi$, i.e., $M(\pi)=M(\minf(\pi))$. Run $\pi$ is \emph{accepting} iff
$M(\pi)$ satisfies the formula $\varphi$, where $M(\pi)$ satisfies
$\Inf m$ iff $m\in M(\pi)$ and $M(\pi)$ satisfies $\Fin m$ iff
$m\not\in M(\pi)$. The \emph{language} of automaton $\mA$ is the set
$L(\mA)=\{u\in \Sigma^{\omega}\mid\textrm{there is an accepting run of
}\mA\textrm{ over }u\}$. Two automata $\mA,\mB$ are \emph{equivalent}
if $L(\mA)=L(\mB)$.
% such that $\mA$ has
% an accepting run over $u$.

@@

A \emph{path} from a state $p$ to a state $q$ is a finite sequence of
adjacent transitions of the form
$(q_0,u_0,M_0,q_1)(q_1,u_1,M_1,q_2)\dots(q_{n-1},u_{n-1},M_{n-1},q_n)\in\delta^+$
such that $p=q_0$ and $q=q_n$. We assume that each state $q$ is
reachable from the initial state $q_I$ (i.e., $q=q_I$ or there is a
path from $q_I$ to $q$) as states that are not reachable from $q_I$
can be eliminated without any impact on $L(\mA)$. A nonempty set of
states $S\subseteq Q$ is called a nontrivial \emph{strongly connected
  component (SCC)} if for each $p,q\in S$ there is a path from $p$ to
$q$. An SCC S is \emph{maximal} if there is no SCC $S'$ satisfying
$S\subsetneq S'$. In the rest of this paper, SCC always refers to a
maximal SCC. Given an SCC $S$,
$\delta_S=\delta\cap (S\times \Sigma \times 2^M \times S)$ denotes the
set of all transitions between states in $S$. Further, for each mark
$m\in M$, let $\delta_m$ denotes the set of all transitions marked
with $m$, i.e., $\delta_m=\{(p,a,M',q)\in\delta\mid m\in M'\}$. The
sets $\delta_m$ are called \emph{acceptance sets}.  Given an SCC $S$
and a mark $m\in M$, by $\delta_{S,m}$ we denote the set
$\delta_{S,m}=\delta_S\cap\delta_m$. Finally, a set of transitions
$C\subseteq\delta$ is called a \emph{cycle} is there exists a path
from a state $p$ to the same state containing all transitions of $C$
and no transition outside $C$.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,q_I,\delta,\varphi)$ without altering its
structure and language. In other words, we change only the acceptance
marks on individual transitions and we replace the acceptance formula
by a new one with less acceptance marks. To ensure that the modified
automaton is equivalent to the original one, we actually construct the
new acceptance formula such that the set of acceptance runs stays the
same (modulo changed acceptance marks on transitions).

We start with the several simple observations. Acceptance of a run
$\pi$ is fully determined by the set $M(\pi)$ of acceptance marks that
appear infinitely often on $\pi$, which is in turn fully determined by
the set $\inf(\pi)$ of transitions appearing infinitely often in
$\pi$.
% \[
%   M(\pi)=\bigcup_{(q_i,u_i,M_i,q_{i+1})\in\minf(\pi)}M_i
% \]
Further, each set of the form $\inf(\pi)$ is a cycle and vice versa.
Hence, our reduction methods is looking for

following simple observation. Definovat $inf(\varphi)$?
